---
title: "WGS code - part 1 (Manhattan + Stats)"
author: "Tyler Weide"
date: "October 20th, 2025"
format:
  html:
    theme: "Cosmo"
    toc: true
    number-depth: 3
    html-math-method: katex
    code-fold: true
    page-layout: full
    embed-resources: true
    standalone: true
---

# Setup Knitr options

```{r}
#| label: knitr_setup
#| warning: false
#| echo: false

# set options for knitr (engine that will render everything)
knitr::opts_chunk$set(echo       = TRUE,
                      warning    = FALSE,
                      error      = FALSE,
                      message = FALSE,
                      fig.height = 6,
                      fig.width  = 8
                      )

```

# Load in the Files
```{r}
#| label: setup_code 
#| warning: FALSE
#| echo: TRUE

#==============================================================================#
# Setup Options
#==============================================================================#

# remove all objects if restarting script

rm(list=ls())

# set options for R
options(
  tibble.width = Inf,   # print all columns in tibble
  scipen = 999          # remove scientific notation
  )

suppressPackageStartupMessages(library(tidyverse))
#==============================================================================#
# Packages
#==============================================================================#

# load libraries
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(qqman)
library(data.table)
library(arrow)
library(qvalue)
#==============================================================================#

#==============================================================================#

# set all paths
path_main    <- "."
#path_data    <- str_c(path_main, "Data/",    sep="")
#path_plots   <- str_c(path_main, "Plots",   sep="")
#path_scripts <- str_c(path_main, "Scripts/", sep="")

#==============================================================================#
# Set Inputs
#==============================================================================#


# Function to load in the files (that have THE SAME column names)
read_pop <- function(path) {
  read_table(
    file = path,
    skip = 1,
    col_names = c("CHROM","POS","ALT_DP","REF_DP","DP_TOTAL"),
    col_types = "cnnnn")
}

# These are the files that we are loading in with the above function
B1_low    <- read_pop("B1_low_AF_input.csv")
B1_high   <- read_pop("B1_high_AF_input.csv")
B1_merged <- read_pop("B1_merged_AF_input.csv")

B2_low    <- read_pop("B2_low_AF_input.csv")
B2_high   <- read_pop("B2_high_AF_input.csv")
B2_merged <- read_pop("B2_merged_AF_input.csv")

B3_low    <- read_pop("B3_low_AF_input.csv")
B3_high   <- read_pop("B3_high_AF_input.csv")
B3_merged <- read_pop("B3_merged_AF_input.csv")

# This is the reference map that has the contig information --> chromosome number
refmap <- read_delim(
  file = "refseq_to_chr.tsv",
  delim = "\t",
  col_names = c("CHROM","CHR"),
  col_types = "cc")

```


# Setting up the data after loading in
## Merge the populations into 1 file per boar (Low, High, Merged)
```{r}
#| label: Merging populations into 1 file per boar
#| warning: FALSE
#| echo: TRUE

## Merging the low, high, merged populations into 1 file, per boar 

# Boar 1
B1_all <- B1_low %>%
  full_join(B1_high,  by = c("CHROM", "POS"), suffix = c("_low", "_high")) %>%
  full_join(B1_merged, by = c("CHROM", "POS")) %>%
  rename_with(~ paste0(., "_merged"), c("ALT_DP", "REF_DP", "DP_TOTAL"))

# Boar 2
B2_all <- B2_low %>%
  full_join(B2_high,  by = c("CHROM", "POS"), suffix = c("_low", "_high")) %>%
  full_join(B2_merged, by = c("CHROM", "POS")) %>%
  rename_with(~ paste0(., "_merged"), c("ALT_DP", "REF_DP", "DP_TOTAL"))

# Boar 3
B3_all <- B3_low %>%
  full_join(B3_high,  by = c("CHROM", "POS"), suffix = c("_low", "_high")) %>%
  full_join(B3_merged, by = c("CHROM", "POS")) %>%
  rename_with(~ paste0(., "_merged"), c("ALT_DP", "REF_DP", "DP_TOTAL"))

```

## Remove all 0's in either ALT_DP or REF_DP
```{r}
#| label: Removing 0's from ALT or REF reads
#| warning: FALSE
#| echo: TRUE

# Function to count removed SNPs
count_removed_snps <- function(before, after, boar_name) {
  removed <- nrow(before) - nrow(after)
  cat(boar_name, ":\n",
      "  Original SNPs:", nrow(before), "\n",
      "  Remaining SNPs:", nrow(after), "\n",
      "  Removed SNPs:", removed, "\n\n")
}

# Before filtering (save copies)
B1_before <- B1_all
B2_before <- B2_all
B3_before <- B3_all

# Apply filtering to get rid of any SNP that has 0's in either ALT or REF Depth
filter_zero_dp <- function(df) {
  df %>%
    filter(
      ALT_DP_low    != 0,
      REF_DP_low    != 0,
      ALT_DP_high   != 0,
      REF_DP_high   != 0,
      ALT_DP_merged != 0,
      REF_DP_merged != 0)
}

# Applying the above function to all of the populations)
B1_all <- filter_zero_dp(B1_before)
B2_all <- filter_zero_dp(B2_before)
B3_all <- filter_zero_dp(B3_before)

# Count and print summary
count_removed_snps(B1_before, B1_all, "Boar 1")
count_removed_snps(B2_before, B2_all, "Boar 2")
count_removed_snps(B3_before, B3_all, "Boar 3")

```

## Calculating allele frequencies to each population per boar
```{r}
#| label: Calculating AF
#| warning: FALSE
#| echo: TRUE

# Function that adds AF for each boar for the ALT AF
add_AF <- function(df) {
  df %>%
    mutate(
      AF_low    = ALT_DP_low    / (ALT_DP_low    + REF_DP_low),
      AF_high   = ALT_DP_high   / (ALT_DP_high   + REF_DP_high),
      AF_merged = ALT_DP_merged / (ALT_DP_merged + REF_DP_merged))
}

# Using the above function on all boars
B1_all <- add_AF(B1_all)
B2_all <- add_AF(B2_all)
B3_all <- add_AF(B3_all)
```

## Filter SNPs that only have a heterozygous merged population (Merged_AF between 0.4-0.6) and add Chromosome number to each SNP
```{r}
#| label: Filter for heterozygous merged population
#| warning: FALSE
#| echo: TRUE

# function to filter for heterozygous SNPs in merged population and add Chromosome number to B*_All nect to Contig info
# Prints out how many SNPs are removed and kept, and saves them as a .csv
process_boar <- function(df, name) {
  filtered <- df %>%
    filter(AF_merged >= 0.4 & AF_merged <= 0.6) %>%
    left_join(refmap, by = "CHROM") %>%
    relocate(CHR, .after = CHROM)
  
  cat(name, "- SNPs kept:", nrow(filtered),
      "| SNPs removed:", nrow(df) - nrow(filtered), "\n")
  
  write.csv(filtered, paste0(name, "_QC_SNPs.csv"), row.names = FALSE)
  return(filtered)
}

# Runs above function
B1_all_filtered <- process_boar(B1_all, "Boar1")
B2_all_filtered <- process_boar(B2_all, "Boar2")
B3_all_filtered <- process_boar(B3_all, "Boar3")
```

## Filter for allele counts in Ref and Alt for all boars and all populations
```{r}
#| label: Filter for allele counts
#| warning: FALSE
#| echo: TRUE

# function to filter for Allele counts, how many SNPs are kept/removed, and write a .csv of those SNPs per boar
filter_dp <- function(df, name) {
  out <- df %>%
    dplyr::filter(
      DP_TOTAL_low >= 30,
      DP_TOTAL_high >= 30,
      ALT_DP_low > 10,  REF_DP_low > 10,
      ALT_DP_high > 10, REF_DP_high > 10,
      ALT_DP_merged > 10, REF_DP_merged > 10)
  cat(name, "- kept:", nrow(out), "| removed:", nrow(df) - nrow(out), "\n")
  write.csv(out, paste0(name, "_QC_SNPs_DP.csv"), row.names = FALSE)
  out
}

# Runs the above function
B1_all_filtered <- filter_dp(B1_all_filtered, "Boar1")
B2_all_filtered <- filter_dp(B2_all_filtered, "Boar2")
B3_all_filtered <- filter_dp(B3_all_filtered, "Boar3")

```

# Stats and Plotting after filtering
## Z test
```{r}
#| label: Binomial, Z test, Manhattan
#| warning: FALSE
#| echo: TRUE

# Combine all boars into 1 object
all_boars <- bind_rows(
  B1_all_filtered %>% mutate(Boar = "Boar1"),
  B2_all_filtered %>% mutate(Boar = "Boar2"),
  B3_all_filtered %>% mutate(Boar = "Boar3"))


# Summarize read counts for each SNP across all three boars
# Each SNP (CHROM + POS) now has total ALT and REF counts in Low and High populations.
counts <- all_boars %>%
  group_by(CHROM, POS) %>%
  summarise(
    ALT_low = sum(ALT_DP_low),
    REF_low = sum(REF_DP_low),
    ALT_high = sum(ALT_DP_high),
    REF_high = sum(REF_DP_high),
    .groups = "drop") %>%
  filter(ALT_low + REF_low > 0, ALT_high + REF_high > 0)

# Compute the total read depth per population for each SNP
n1 <- counts$ALT_low + counts$REF_low
n2 <- counts$ALT_high + counts$REF_high

# Calculate allele frequencies (ALT/total) for each population
p1 <- counts$ALT_low / n1   
p2 <- counts$ALT_high / n2  

# Compute the pooled allele frequency across both groups
p_pool <- (counts$ALT_low + counts$ALT_high) / (n1 + n2)

# Compute the standard error for the pooled 2 proportion Z test
se <- sqrt(p_pool * (1 - p_pool) * (1/n1 + 1/n2))

# Compute the z-statistic for each SNP (difference / standard error)
z <- (p1 - p2) / se

# Convert the z-scores into two-sided p-values
p_value <- 2 * pnorm(-abs(z))


# Create a new data frame containing allele frequencies, delta_AF, and p/q-values + adding in z statistic
ARD_all <- counts %>%
  mutate(
    AF_low = p1,
    AF_high = p2,
    delta_AF = p2 - p1,
    z_stat = z,
    p_value = p_value,
    q_value = p.adjust(p_value, method = "BH"),
    neglogP = -log10(p_value))

# Adding CHR number next to contig information for each SNP, adding back SNP information
ARD_all <- ARD_all %>%
  left_join(refmap, by = "CHROM") %>%
  relocate(CHR, .after = CHROM) 


```

## Manhattan Plot
```{r}
#| label: Manhattan Plot, Histogram, Log Uniform Q-Q 
#| warning: FALSE
#| echo: TRUE

# Defines chromosome order
chr_levels <- c(as.character(1:18), "X", "Y")

# Make sure chromosomes and plotting order is correct and makes sure every snp has a -log10(p) value, if it doesn't, it will calculate
man_df <- ARD_all %>%
  filter(CHR %in% chr_levels) %>%
  mutate(CHR = factor(CHR, levels = chr_levels)) %>%
  arrange(CHR, POS) %>%
  mutate(neglogP = ifelse(is.na(neglogP), -log10(p_value), neglogP))


# Chromosome table that calculates lengths and offsets of the manhattan plot, where 1 ends, the other begins e.g, 3 = length of 2 + 1
chr_tbl <- man_df %>%
  group_by(CHR) %>%
  summarise(chr_len = max(POS, na.rm = TRUE), .groups = "drop") %>%
  mutate(tot = cumsum(chr_len) - chr_len)

# Merges the offsets from above and computes cumulative base-pair position so they will all align continuously on x-axis
man_df <- man_df %>%
  left_join(chr_tbl, by = "CHR") %>%
  mutate(BP_cum = POS + tot,
         colgrp = as.integer(CHR) %% 2)

# Computes chromosome centers for labeling of the X - axis (1:18, X, Y)
axis_df <- chr_tbl %>%
  mutate(center = tot + chr_len/2)

# Set cutoff to -log10(p) = 4 (equivalent to p = 0.0001)
cutoff_line <- 4

# Plotting the actual Manhattan plot
manhattan <- ggplot(man_df, aes(BP_cum, neglogP, color = factor(colgrp))) +
  geom_point(size = 0.28, alpha = 0.85) +
  scale_color_manual(values = c("black", "grey70")) +
  scale_x_continuous(breaks = axis_df$center, labels = axis_df$CHR, expand = c(0,0)) +
  scale_y_continuous(limits = c(0, 10)) +
  geom_hline(yintercept = cutoff_line, color = "red3", linewidth = 0.7, alpha = 0.8) +
  labs(title = "Allelic Ratio Distortion across all three Boars",
       x = "Chromosome", y = expression(-log[10](p))) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5))

manhattan

# Saves the plot 
ggsave("Manhattan_ARD_all3boars.png", manhattan, width = 10, height = 6, units = "in", dpi = 300)

# Counts the number of significant snps above the corrected line of 0.0001
sig_snps <- man_df %>% filter(neglogP > 4)
n_sig <- nrow(sig_snps)
n_sig

```

### Histograms of P-values and Log Uniform Q-Q plot of Manhattan
```{r}
#| label: Histogram and Log Uniform Q-Q plots
#| warning: FALSE
#| echo: TRUE

# Histogram of P values of the manhattan plot
p_hist <- man_df %>%
  mutate(p = 10^(-neglogP)) %>%
  filter(is.finite(p), !is.na(p), p > 0, p <= 1) %>%
  ggplot(aes(x = p)) +
  geom_histogram(bins = 40, boundary = 0, closed = "left",
                 color = "white", fill = "grey40") +
  labs(
    title = "P-value distribution (robust Z test)",
    x = "p-value",
    y = "Count") +
  theme_minimal(base_size = 12)

# Create object from man_df that has 10^(-neglogP) and filter for NAs
p_longer <- man_df %>%
  mutate(p = 10^(-neglogP)) %>%
  filter(is.finite(p), !is.na(p), p > 0, p <= 1) %>%
  mutate(nlogp = -log10(p))

# Create Rank, expected vs. observed for QQ plot
qq_df <- p_longer %>%
  arrange(p) %>%
  mutate(
    rank = row_number(),
    exp_u = (rank - 0.5) / n(),
    exp_log = -log10(exp_u),
    obs_log = nlogp)

# Plotting QQ plot
p_qq <- ggplot(qq_df, aes(x = exp_log, y = obs_log)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(size = 0.5, alpha = 0.5) +
  labs(
    title = "Q–Q plot against log-uniform null",
    x = "Expected -log10(p) Uniform[0,1]",
    y = "Observed -log10(p)") +
  theme_minimal(base_size = 12)

z_test <- ggplot(ARD_all, aes(x = z_stat)) +
  geom_histogram(aes(y = ..density..), bins = 100, alpha = 0.6) +
  stat_function(fun = dnorm, color = "red", size = 1) +
  xlab("Z statistic") +
  ggtitle("Z distribution with standard normal overlay")

# Printing out the plots
z_test
p_hist
p_qq

# Save plots
ggsave("Histogram of Z-test.png", z_test, width = 10, height = 6, dpi = 300)
ggsave("Histogram of Manhattan plot P-values.png", p_hist, width = 10, height = 6, dpi = 300)
ggsave("Q-Q plot of Manhattan plot P-values.png", p_qq, width = 10, height = 6, dpi = 300)


# Calculate genomic inflation (are the observed test statistics more extreme than expected under the null? Should be ~ 1)
lambda_gc <- function(p){
  p <- p[is.finite(p)]
  p <- pmin(pmax(p, 1e-300), 1 - 1e-16)
  chisq <- qchisq(1 - p, 1)
  median(chisq, na.rm = TRUE) / qchisq(0.5, 1)
}

# Print out value, anything over 1.20 is not good
lambda_value <- lambda_gc(10^(-man_df$neglogP))
lambda_value
```


## Ideogram style plot
```{r}
#| label: Ideogram plot
#| warning: FALSE
#| echo: TRUE

# Right order, could have been called using previous chr_levels vector
chr_order <- c(as.character(1:18), "X", "Y")

# spacing controls for the plot
spacing    <- 1.6  
bar_height <- 0.7   

# Chromosome lengths and y positions
chr_tbl <- man_df %>%
  filter(CHR %in% chr_order) %>%
  group_by(CHR) %>%
  summarise(chr_len = max(POS, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    ord   = match(CHR, chr_order),
    y_pos = (length(chr_order) - ord + 1) * spacing,          
    xmin  = 0,
    xmax  = chr_len / 1e6,
    ymin  = y_pos - bar_height/2,
    ymax  = y_pos + bar_height/2)

# Significant SNPs at same y positions
sig_pts <- man_df %>%
  filter(CHR %in% chr_order, neglogP > 4) %>%                 
  select(CHR, POS) %>%
  left_join(chr_tbl %>% select(CHR, y_pos), by = "CHR") %>%
  mutate(xMb = POS / 1e6)

# Plotting the ideogram style map
Ideogram_plot <- ggplot() +
  geom_rect(data = chr_tbl,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey88", color = NA) +
  geom_point(data = sig_pts,
           aes(x = xMb, y = y_pos),
           size = 1.2,      
           alpha = 0.8,
           color = "#1F78B4") +
  scale_y_continuous(
    breaks = chr_tbl$y_pos,
    labels = paste0("chr", chr_tbl$CHR),
    expand = expansion(mult = c(0.02, 0.06))) +
  labs(title = "Significant SNP positions",
       x = "Position (Mb)", y = NULL) +
  theme_classic(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y  = element_text(face = "bold"),
    plot.title   = element_text(face = "bold", hjust = 0.5))

Ideogram_plot

ggsave("Significant_SNP_Ideogram.png", Ideogram_plot, width = 10, height = 6, dpi = 300)

```
# Pairing up SNPs with boar for identification
## Find which SNPs map to which BOAR, CHROM, POS, and which SNPs are shared between
```{r}
#| label: Matching SNPs
#| warning: FALSE
#| echo: TRUE


# Create list of significant SNPs from manhattan plot
sig <- sig_snps %>% distinct(CHROM, POS, .keep_all = TRUE)

# Filter these snps only by CHROM and POS
b1_keys <- B1_all_filtered %>% distinct(CHROM, POS) %>% mutate(B1 = TRUE)
b2_keys <- B2_all_filtered %>% distinct(CHROM, POS) %>% mutate(B2 = TRUE)
b3_keys <- B3_all_filtered %>% distinct(CHROM, POS) %>% mutate(B3 = TRUE)

# Merge these counts together, and get counts of how many SNPs are shared between each boar
sig_boars <- sig %>%
  left_join(b1_keys, by = c("CHROM","POS")) %>%
  left_join(b2_keys, by = c("CHROM","POS")) %>%
  left_join(b3_keys, by = c("CHROM","POS")) %>%
  mutate(across(c(B1,B2,B3), ~replace_na(.x, FALSE)),
         boar_count = B1 + B2 + B3,
         boar_label = case_when(
           boar_count == 3 ~ "B1,B2,B3",
           boar_count == 2 & B1 & B2 ~ "B1,B2",
           boar_count == 2 & B1 & B3 ~ "B1,B3",
           boar_count == 2 & B2 & B3 ~ "B2,B3",
           boar_count == 1 & B1 ~ "B1",
           boar_count == 1 & B2 ~ "B2",
           boar_count == 1 & B3 ~ "B3",
           TRUE ~ "None"))

# Summarize these counts
sig_counts <- sig_boars %>% count(boar_label, sort = TRUE)

# Make a .csv file with the SNP information for all significant SNPs
write_csv(sig_boars, "ARD_sig379_with_boars.csv")
write_csv(sig_counts, "ARD_sig379_counts_by_boar.csv")

## Split the data by which boars share each SNP, Contains all SNP info

# One boar only
b1_snps <- sig_boars %>% filter(boar_label == "B1")
b2_snps <- sig_boars %>% filter(boar_label == "B2")
b3_snps <- sig_boars %>% filter(boar_label == "B3")

# Shared between two
b1b2_snps <- sig_boars %>% filter(boar_label == "B1,B2")
b1b3_snps <- sig_boars %>% filter(boar_label == "B1,B3")
b2b3_snps <- sig_boars %>% filter(boar_label == "B2,B3")

# Shared among all three
shared_all <- sig_boars %>% filter(boar_label == "B1,B2,B3")

# Write each to CSV
write_csv(b1_snps, "B1_unique_SNPs.csv")
write_csv(b2_snps, "B2_unique_SNPs.csv")
write_csv(b3_snps, "B3_unique_SNPs.csv")
write_csv(shared_all, "Shared_all3_SNPs.csv")

```

### Write CSVs for only CHROM and POS for the above chunk
```{r}
#| label: Write CSVs for CHROM and POS
#| warning: FALSE
#| echo: TRUE

## Same as above chunk but only saving the CHROM and POS information for downstream analysis (SNPeff)

# One boar only
b1_snps_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B1") %>% select(CHROM, POS)
b2_snps_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B2") %>% select(CHROM, POS)
b3_snps_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B3") %>% select(CHROM, POS)

# Shared between two
b1b2_snps_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B1,B2") %>% select(CHROM, POS)
b1b3_snps_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B1,B3") %>% select(CHROM, POS)
b2b3_snps_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B2,B3") %>% select(CHROM, POS)

# Shared among all three
shared_all_CHROMPOSonly <- sig_boars %>% filter(boar_label == "B1,B2,B3") %>% select(CHROM, POS)

# Write each to CSV
write_csv(b1_snps_CHROMPOSonly, "SNPs_B1_unique.csv")
write_csv(b2_snps_CHROMPOSonly, "SNPs_B2_unique.csv")
write_csv(b3_snps_CHROMPOSonly, "SNPs_B3_unique.csv")
write_csv(b1b2_snps_CHROMPOSonly, "SNPs_B1B2_shared.csv")
write_csv(b1b3_snps_CHROMPOSonly, "SNPs_B1B3_shared.csv")
write_csv(b2b3_snps_CHROMPOSonly, "SNPs_B2B3_shared.csv")
write_csv(shared_all_CHROMPOSonly, "SNPs_all3_shared.csv")


```

# Fishers Exact Test for each Sig SNP per boar 
```{r}
#| label: Fishers Exact Test per boar
#| warning: FALSE
#| echo: TRUE

# This will create a function that creates/joins the significant SNPs into a table with all relevant information and perform a FET on a per boar basis for each SNP

# This tests whether or not the SNPs that were above the Log(pVal) = 4 on the manhattan are actually different in a single boar rather than combined

add_boar_fisher <- function(sig_df, boar_df, boar_name) {
  boar_counts <- boar_df %>%
    select(CHROM, POS, ALT_DP_low, REF_DP_low, ALT_DP_high, REF_DP_high) %>%
    rename(
      ALT_low  = ALT_DP_low,
      REF_low  = REF_DP_low,
      ALT_high = ALT_DP_high,
      REF_high = REF_DP_high) %>%
    rename_with(~ paste0(., "_", boar_name),
                c(ALT_low, REF_low, ALT_high, REF_high))

  sig_df %>%
    left_join(boar_counts, by = c("CHROM", "POS")) %>%
    mutate(
      !!paste0("fisher_p_", boar_name) :=
        pmap_dbl(
          list(
            .data[[paste0("ALT_low_",  boar_name)]],
            .data[[paste0("REF_low_",  boar_name)]],
            .data[[paste0("ALT_high_", boar_name)]],
            .data[[paste0("REF_high_", boar_name)]]),
          function(ALT_low, REF_low, ALT_high, REF_high) {
            if (any(is.na(c(ALT_low, REF_low, ALT_high, REF_high)))) return(NA_real_)
            mat <- matrix(c(ALT_high, REF_high, ALT_low, REF_low),
                          nrow = 2, byrow = TRUE)
            if (sum(mat) == 0) return(NA_real_)
            fisher.test(mat)$p.value
          }
        )
    )
}

sig_fisher <- sig_boars %>%
  add_boar_fisher(B1_all_filtered, "B1") %>%
  add_boar_fisher(B2_all_filtered, "B2") %>%
  add_boar_fisher(B3_all_filtered, "B3")

```

### Table of FET p values
```{r}
#| label: Table of FET 
#| warning: FALSE
#| echo: TRUE

# This creates a more tidy table that has all significant SNPs and associated boars with FET p values
sig_table <- sig_fisher %>%
  select(
    CHROM, CHR, POS, boar_label,
    starts_with("ALT_low_"),
    starts_with("REF_low_"),
    starts_with("ALT_high_"),
    starts_with("REF_high_"),
    starts_with("fisher_p_")) %>%
  pivot_longer(
    cols = -c(CHROM, CHR, POS, boar_label),
    names_to = c("metric", "boar"),
    names_pattern = "(.+)_((?:B1|B2|B3))",
    values_to = "value") %>%
  pivot_wider(
    names_from = metric,
    values_from = value) %>%
  filter(!(is.na(ALT_low) & is.na(REF_low) & is.na(ALT_high) & is.na(REF_high))) %>%
  mutate(
    AF_low  = ALT_low  / (ALT_low  + REF_low),
    AF_high = ALT_high / (ALT_high + REF_high)) %>%
  arrange(CHR, POS, boar)

view(sig_table)

```

### Filtering out FET p values
```{r}
#| label: Filtering out FET p values
#| warning: FALSE
#| echo: TRUE

# Filter out p values less than 0.05 from FET
filtered_sig <- sig_table %>%
  filter(fisher_p < 0.050)

# Counting the number of total rows, kept rows, and removed rows
total_rows   <- nrow(sig_table)
kept_rows    <- nrow(filtered_sig)
removed_rows <- total_rows - kept_rows

# Identify unique SNPs that were removed (if any boar fails, most likely from overinflation of AF from Z test)
removed_snps <- sig_table %>%
  filter(fisher_p >= 0.050) %>%
  distinct(CHROM, POS)

n_removed_snps <- nrow(removed_snps)

# Count remaining SNPs after filtering
kept_snps <- filtered_sig %>%
  distinct(CHROM, POS)

n_kept_snps <- nrow(kept_snps)

# Count how many SNP–boar pairs passed FET per boar
kept_per_boar <- filtered_sig %>%
  count(boar)

# Count SNPs fully supported in all boars they appear in
fully_supported <- filtered_sig %>%
  group_by(CHROM, POS) %>%
  summarise(
    n_boars_tested = n(),
    n_boars_passed = sum(fisher_p < 0.05),
    .groups = "drop") %>%
  filter(n_boars_tested == n_boars_passed)

n_fully_supported <- nrow(fully_supported)

# Summary table of all of the above metrics
summary_table <- tibble(
  metric = c(
    "Total SNP–boar rows",
    "Kept rows (p < 0.05)",
    "Removed rows (p ≥ 0.05)",
    "Unique SNPs removed",
    "Unique SNPs kept",
    "Fully supported SNPs"),
  count = c(
    total_rows,
    kept_rows,
    removed_rows,
    n_removed_snps,
    n_kept_snps,
    n_fully_supported))

summary_table



```
### Filtering and Counting Significant SNPs 
```{r}
#| label: Filtering and counting Sig SNPs
#| warning: FALSE
#| echo: TRUE

# Find out how many SNPs belong to each boar after filtering
# B1, B2, B3, B1B2, B1B3, B2B3 might inflate due to the removal of boars in combined SNPs, this is okay and expected

# Filter FET p-values from Significant SNP
filtered_sig <- sig_table %>%
  filter(fisher_p < 0.050)

# Counts of SNPs per boar
boar_combination_counts <- filtered_sig %>%
  group_by(CHROM, POS) %>%
  summarise(
    boar_label = paste(sort(unique(boar)), collapse = ","),
    .groups = "drop") %>%
  count(boar_label, sort = TRUE)

# Adding a total row
total_row <- tibble(
  boar_label = "Total significant SNPs remaining",
  n = sum(boar_combination_counts$n))

# Binding the total row with the count table
boar_combination_counts_with_total <- bind_rows(
  boar_combination_counts,
  total_row)

boar_combination_counts_with_total

# Save filtered Sig SNPs as .csv for GO use later
write.csv(filtered_sig, "filtered_sig.csv", row.names = FALSE)

write.csv(
  boar_combination_counts_with_total,
  "boar_combination_counts_with_total.csv",
  row.names = FALSE)

```


# Plotting allele distributions for High/Low and Merged populations per boar
```{r}
#| label: Plotting AF distributions
#| warning: FALSE
#| echo: TRUE

# Join B1, B2, B3 from Significant SNPs with filters
sig_snps_filtered <- filtered_sig %>%
  distinct(CHROM, POS)

B1_sig <- B1_all_filtered %>%
  semi_join(sig_snps_filtered, by = c("CHROM", "POS"))

B2_sig <- B2_all_filtered %>%
  semi_join(sig_snps_filtered, by = c("CHROM", "POS"))

B3_sig <- B3_all_filtered %>%
  semi_join(sig_snps_filtered, by = c("CHROM", "POS"))

# Bind rows together into 1 table, create High/Low labels
af_long_sig <- bind_rows(
  B1_sig %>% transmute(boar = "B1", AF_low, AF_high),
  B2_sig %>% transmute(boar = "B2", AF_low, AF_high),
  B3_sig %>% transmute(boar = "B3", AF_low, AF_high)) %>%
  pivot_longer(cols = c(AF_low, AF_high),
               names_to = "population",
               values_to = "AF") %>%
  filter(!is.na(AF), AF > 0, AF < 1) %>%
  mutate(population = factor(
    population,
    levels = c("AF_low", "AF_high"),
    labels = c("Low", "High")))

# Create plots for High/Low AF populations
p_af_sig <- ggplot(af_long_sig, aes(x = AF, fill = population, colour = population)) +
  geom_density(alpha = 0.35, adjust = 0.4) +
  facet_grid(~ boar) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_fill_manual(values = c("Low" = "#00B0F6",
                               "High" = "#F8766D")) +
  scale_colour_manual(values = c("Low" = "#00B0F6",
                                 "High" = "#F8766D")) +
  labs(
    title = "Allelic Ratio Distributions of High and Low populations per boar",
    x = "ALT Allelic Ratio",
    y = "Density",
    fill = "",
    colour = "") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    panel.spacing = unit(1.5, "lines"))

# Bind rows together for the merged
af_merged_sig <- bind_rows(
  B1_sig %>% transmute(boar = "B1", AF = AF_merged),
  B2_sig %>% transmute(boar = "B2", AF = AF_merged),
  B3_sig %>% transmute(boar = "B3", AF = AF_merged)) %>%
  filter(!is.na(AF), AF > 0, AF < 1)

# Plot merged AF 
p_af_merged_sig <- ggplot(af_merged_sig, aes(x = AF)) +
  geom_density(fill = "#00B0F6", colour = "#00B0F6", alpha = 0.35, adjust = 2) +
  facet_grid(~ boar, labeller = labeller(boar = c(B1 = "S1", B2 = "S2", B3 = "S3"))) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    title = "Allele Frequency Distribution of Merged population per boar",
    x = "Reference Allelic Ratio",
    y = "Density") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    panel.spacing = unit(1.5, "lines"))

# Print plots
p_af_sig
p_af_merged_sig

# Save the plots
ggsave("AFDistributions.png", p_af_sig, width = 10, height = 6, dpi = 300)
ggsave("MergedAFDistributions.png", p_af_merged_sig, width = 10, height = 6, dpi = 300)


```
